---
title: "Minimal example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Minimal example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

## Getting started

### Installation
Install the release version from CRAN:

```{r, eval = FALSE}
install.packages("spectre")
```

To install the developmental version of `spectre`, use:
```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("r-spatialecology/spectre") #Uncomment when repo is public
```

## Example

```{r}
library(spectre)
```

For this example, we created a random species composition (15 sites, gamma diversity = 20) and calculated $\alpha$-diversity and Bray-Curtis dissimilarity to be used as input for the `spectre` algorithm (please see `R/generate_minimal_example_data.R` for details). 

```{r}
# load "observed" alpha-, beta- and gamma-diversity values of the random species composition
alpha_list <- minimal_example_data$alpha_list # richness
beta_list <- minimal_example_data$beta_list # Bray-Curtis dissimilarity
total_gamma <- dim(minimal_example_data$species_list)[1] # 20 species
```

### Solve species composition using `spectre`

Running the optimization is straightforward in `spectre` and only includes one function call.
However, first we need to generate to commonness matrix from the `gdm` predictions to generate the object matrix. 
Then, we simply run the optimization using the alpha list, the $gamma$-diversity, the objective matrix (or target), and finally the number of maximum iterations.

```{r, message = FALSE, warning = FALSE}
library("spectre")

objective_matrix <- spectre::generate_commonness_matrix_from_gdm(gdm_predictions = beta_list, 
                                                                 alpha_list = alpha_list)

# Solve composition 
res <- spectre::run_optimization_min_conf(alpha_list = alpha_list,
                                          total_gamma = total_gamma,
                                          target = objective_matrix,
                                          max_iterations = 1000, # n iterations
                                          seed = 123) # use a random seed for reproducibility
```

### Result evaluation

`spectre` allows to easily calculate some error measures, namely the mean absolute commonness error ($MAE_c$) and the relative commonness error [%] (RCE). 
The error is calculated between the solved species composition and the objective matrix.

```{r, message = FALSE}
error_c <- spectre::calc_commonness_error(x = res, objective_matrix = objective_matrix)
```

The *objective matrix* had a mean commonness of `r round(mean(abs(objective_matrix), na.rm = TRUE), 2)`. 
The mean absolute error between the *objective matrix* and the solved *solution matrix* was `r round(error_c[[1]], 2)`. 
The solution matrix had an relative commonness error (RCE) of `r round(error_c[[2]], 1)` %.

There are also two functions to visualize the optimization. 
First you can plot the error of the solution matrix over time. 
Second, you can plot the commonness error between the solution matrix and the objective matrix.

```{r, include = TRUE, out.width="50%", fig.align="center"}
# With an increasing number of iterations, the solution matrix improved
spectre::plot_error(x = res)

# Plot commonness error between objective matrix and solution matrix
spectre::plot_commonness(x = res, target = objective_matrix)
```

--- 

## References
Matthew C. Fitzpatrick, Karel Mokany, Glenn Manion, Matthew Lisk, Simon Ferrier and Diego Nieto-Lugilde (2021). gdm: Generalized Dissimilarity Modeling. R package version 1.4.2.2. https://CRAN.R-project.org/package=gdm